Think deep for reef resilience

### 2025-09-13 paper/paper_draft.qmd
Ensure retained larvae is treated separately, and is not double counted in incoming or outgoing connectivity

Depth characteristics for each reef were extracted using 10m resolution bathymetry data made
available in GBRMPA [-@GBRMPA2020]. While the bathymetry data adopted for the given
assessment are said to be accurate up to 15m from Mean Sea Level [@GBRMPA2020], there are
many sources of uncertainty in the depth estimates used for this study. Notably, the adopted
depth estimates were derived from bathymetries based on satellite imagery. The bathymetry
data can be limited in their alignment with the established GBRMPA reef outlines used to
characterise reef locations and their areal extent. For each reef, bathymetry data were
spatially matched with the 3806 reef outlines through geometric intersections.

A Quality Assessment and Control process identified reefs with potential data limitations:
reefs with no overlapping bathymetry data, reefs with minimum depth values above sea level
(indicating potential data artifacts), and reefs where bathymetry pixels covered less than
5% of the reef polygon area. This process removed 842 reefs, or 22%, from the initial set of
3806 reefs. Furthermore, reefs that are not associated with GBRMPA Management areas
[@GBRMPA2007] or reefal bioregions [@GBRMPA2024] were also removed from the investigation.
Bioregions which contained fewer than three reefs after these processes were also excluded.
As a result, 1377 reefs were excluded (36%), leaving a total of 2429 remaining for the
purpose of this investigation. All depth values were converted to positive values (depths
below sea level). The median depth was used to provide an indicative depth value for each
reef. Further limitations are outlined in the Discussion. Code to extract depth estimates,
and the resulting depth estimates themselves, are available at [@iwanaga2025].

@@

We adapted an exploratory scenario modelling approach for this investigation. Specifically,
we use combinations of model factors to investigate a wide range of potential futures under
different assumptions of environmental (climatic) and ecological conditions. Exploratory
modelling approaches have gained favour in recent years to explore the implications of deep
uncertainty; where there is little or no agreement on the key drivers of a system, the
extent of the relationship, how best to represent them or their probability distributions
[@conradt2024; @lempert2013]. Exploratory modelling embraces uncertainty as an inherent
characteristic of complex environmental systems [@lempert2010; @lempert2013] and is a core
methodology underpinning the development of robust and adaptive management strategies
[@haasnoot2013].

Rather than attempting to forecast precise outcomes, the exploratory approach aims to
systematically search many possible futures to identify patterns that lead to outcomes of
interest. The approach allows for the identification of robust management strategies that
perform well across a diversity of future conditions rather than attempting to identify a
single most optimal strategy. Concurrently, assessing the wide swathe of scenarios allows
contextual understanding of the conditions in which strategies may fail. Multiple
interacting and compounding sources of uncertainty, and a lack of broad agreement as to
their likelihood, create inherently uncertain outcomes. Here, exploratory modelling was
applied to identify the environmental and reef characteristics, such as thermal stress,
depth, and connectivity, that support favourable outcomes in reef recovery and resilience.

@@

Two separate systems were employed to explore scenarios: ADRIA (the Adaptive Dynamic Reef
Intervention Algorithms platform) and CoralBlox. ADRIA is a multi-criteria decision support
platform designed for the assessment and analysis of coral reef intervention strategies
under uncertainty [@iwanaga2023]. The ADRIA platform provides functionality for scenario
generation and analysis as well as providing the technical infrastructure to couple
environmental and ecosystem representations. CoralBlox is a computationally efficient coral
population model. It simulates functional/morphological groups of corals over their life
cycles, competing for available space over time [@ribeirodealmeida2024]. The coupled
framework is referred to as ADRIA-CoralBlox.

The coral demographic representation and associated model implementation are further
detailed in the supplementary material. Of important note is that the representation of
bleaching mortalities followed Bozec et al., [-@bozec2022] albeit with a scaling coefficient
to incorporate the amelioration effect that depth may have, following Baird et al.,
[@baird2018]. This depth-scaling reduces the proportion of coral colonies impacted by high
heat stress as depth increases. The impacts of tropical cyclones, wave activity, Crown of
Thorns Starfish or coral disease were not included in this investigation, in order to focus
on assessing reef resilience under possible future warming scenarios.

ADRIA-CoralBlox was run for 4096 scenarios for each GCM (20,480 total model runs). Here, a
scenario represents a unique combination of assumed environmental (namely one of the five
DHW trajectories) and ecological factors including coral growth and mortality rates, and
coral thermal tolerances. Scenarios were generated using the low-discrepancy Sobol' sampling
method [@sobol1967; @sobol1993] with Owen Scrambling [@owen1995], a randomised Quasi Monte
Carlo approach which balances scenario exploration with improved convergence properties
compared to pseudo-random sampling. A detailed comparative analysis of sampling methods can
be found in [@kucherenko2015]. For the purpose of scenario assessment, the resulting coral
cover time series for each model run were scaled by the total area of each reef to represent
cover as a percentage of the total estimated reef area. The scenario-median timeseries for
each reef was used for further analyses. These time series of coral cover were then
clustered based on their temporal behaviour.

@@

To assess the impact of reef connectivity on modelled outcomes we summarise incoming
connectivity (weighted incoming connectivity), outgoing connectivity (outgoing connection
strength), the ratio of source-to-sink connectivity, and larval retention probability for
each reef. Weighted incoming connectivity is calculated by summing the product of each
connection strength incoming to a target reef and the mean timeseries absolute coral cover
(km^2^) of each matching source reef. This metric provides an approximation of overall
larval flow into a reef, with high values indicating the target reef is supplied high
volumes of larvae via strong connections and low values indicating the target reef has weak
(or no) incoming connections or is supplied by small source populations. A limitation of
weighted incoming connectivity is it does not incorporate the proportion of the source
reef's population that is reproducing and producing larvae, instead taking the total coral
cover area as an indicator of larval productivity. Outgoing connection strength is
calculated as the sum of all outgoing transition probabilities from a target reef. Outgoing
connection strength has a theoretical maximum of one (all larvae reach settlement on sink
reefs), however the large size of the GBR combined with larval mortality considerations
result in outgoing connection strength values much lower than one. The ratio of
source-to-sink connectivity ($O:I$) for a target reef represents the strength of the reef's
outgoing larval flow relative to it's incoming larval flow. This is calculated as:
$O:I = \frac{S_O/N_O}{S_I/N_I}$. Where $S$ represents the sum of connectivity probabilities,
$N$ represents the number of connections, $O$ denotes outgoing connectivity from the target
reef and $I$ denotes incoming connectivity to the target reef. If a reef has no incoming
connections, the $O:I$ equals the $S_O/N_O$ value. Incoming connections in this calculation,
do not directly represent the larval supply gained as the breeding populations from which
connections are coming from are not represented. A source-to-sink ratio greater than 1 (a
value greater than 0 on a log scale) indicates a reef has greater outgoing than incoming
larval flow. Reef larval retention probability indicates the proportion of spawned larvae
from a target reef that remain to settlement.

>>>>

Depth characteristics for each reef were extracted using 10m resolution bathymetry data made
available in GBRMPA [-@GBRMPA2020]. For each reef, bathymetry data were spatially matched
with the 3806 reef outlines through geometric intersections. All depth values were converted
to positive values (depths below sea level). The median depth was used to provide an
indicative depth value for each reef. Code, and the resulting depth estimates, are available
at [@iwanaga2025].

While the bathymetry data adopted for the given assessment are said to be accurate up to 15m
from Mean Sea Level [@GBRMPA2020], there are many sources of uncertainty in the depth
estimates used for this study. These are outlined further in the Discussion. Notably, the
adopted depth estimates were derived from bathymetries based on satellite imagery. The
bathymetry data can be limited in their alignment with the established GBRMPA reef outlines
used to characterise reef locations and their areal extent. A Quality Assessment and Control
process identified reefs with potential data limitations: reefs with no overlapping
bathymetry data, reefs with minimum depth values above sea level (indicating potential data
artifacts), and reefs where bathymetry pixels covered less than 5% of the reef polygon area.
This process removed 842 reefs, or 22%, from the initial set of 3806 reefs. Furthermore,
reefs that are not associated with GBRMPA Management areas [@GBRMPA2007] or reefal
bioregions [@GBRMPA2024] were also removed from the investigation. Bioregions which
contained fewer than three reefs after these processes were also excluded. As a result, 1377
reefs were excluded (36%), leaving a total of 2429 remaining for the purpose of this
investigation.

@@

We adapted exploratory scenario modelling approach from robust decision making (RDM)
approaches for this investigation. In particular, we used parameter sampling and exploratory
modelling to investigate a wide range of plausible futures under different assumptions of
environmental (climatic) conditions and ecological factors. Exploratory modelling approaches
have gained favour in recent years as precise long-term predictions of future conditions are
often impossible in complex ecological systems [@conradt2024; @lempert2013]. RDM embraces
uncertainty as an inherent characteristic of complex environmental systems [@lempert2010;
@lempert2013] and is a core methodology underpinning the development of robust and adaptive
management strategies [@haasnoot2013]. Rather than attempting to forecast specific outcomes,
RDM aims to systematically explore many possible futures to identify patterns. Allowing for
the identification of robust management strategies that perform well across those diverse
conditions, as well as conditions in which such strategies may fail. Multiple interacting
and compounding sources of uncertainty, and a lack of broad agreement as to their
likelihood, create inherently uncertain outcomes. While we did not directly model management
decision making in this investigation, we used exploratory modelling to identify the
environmental and reef characteristics, such as thermal stress, depth and connectivity, that
support favourable outcomes such as recovery and resilience.

@@

To apply exploratory scenario modelling, we employed a framework involving two separate
systems: ADRIA (the Adaptive Dynamic Reef Intervention Algorithms platform) and CoralBlox.
ADRIA is a multi-criteria decision support platform designed to help reef managers address
climate change challenges in coral reef systems [@iwanaga2023]. The ADRIA platform provides
functionality for scenario generation and analysis as well as providing the technical
infrastructure to couple environmental and ecosystem representations. CoralBlox is a
computationally efficient coral population model. It simulates functional/morphological
groups of corals over their life cycles, competing for available space over time
[@ribeirodealmeida2024]. The coupled framework is referred to as ADRIA-CoralBlox.

The coral demographic representation and associated model implementation are further
detailed in the supplementary material. Of important note, however, is that the
representation of bleaching mortalities followed Bozec et al., [-@bozec2022] albeit with a
scaling coefficient to incorporate the amelioration effect that depth may have, following
Baird et al., [@baird2018]. This depth-scaling reduces the proportion of coral colonies
impacted by high heat stress as depth increases. The impacts of tropical cyclones, wave
activity, Crown of Thorns Starfish or coral disease were not included in this investigation,
in order to focus on assessing reef resilience under possible future warming scenarios.

ADRIA-CoralBlox was run for 4096 scenarios for each GCM (20,480 total model runs). Here, a
scenario represents a unique combination of assumed environmental (namely one of the five
DHW trajectories) and ecological factors including coral growth and mortality rates, and
coral thermal tolerances. Scenarios were generated using the low-discrepancy Sobol' sampling
method [@sobol1967; @sobol1993] with Owen Scrambling [@owen1995], a random-Quasi Monte Carlo
approach which balances scenario exploration with improved convergence properties compared
to pseudo-random sampling. A detailed comparative analysis of the sampling methods can be
found in [@kucherenko2015]. For the purpose of scenario assessment, the resulting coral
cover time series for each model run were scaled by the total area of each reef to represent
cover as a percentage of the total estimated reef area. The scenario-median timeseries for
each reef was used for further analyses. These time series of coral cover were then
clustered based on their temporal behaviour.

@@

To assess the impact of reef connectivity on modelled outcomes we summarise incoming
connectivity (weighted incoming connectivity), outgoing connectivity (outgoing connection
strength), the ratio of source-to-sink connectivity, and larval retention probability for
each reef. Weighted incoming connectivity is calculated by summing the product of each
connection strength incoming to a target reef (excluding retained larvae from the target
reef) and the mean timeseries absolute coral cover (km^2^) of each matching source reef.
This metric provides an approximation of overall larval flow into a reef, with high values
indicating the target reef is supplied high volumes of larvae via strong connections and low
values indicating the target reef has weak (or no) incoming connections or is supplied by
small source populations. A limitation of weighted incoming connectivity is it does not
incorporate the proportion of the source reef's population that is reproducing and producing
larvae, instead taking the total coral cover area as an indicator of larval productivity.
Outgoing connection strength is calculated as the sum of all outgoing transition
probabilities from a target reef. Outgoing connection strength has a theoretical maximum of
one (all larvae reach settlement on sink reefs), however the large size of the GBR combined
with larval mortality considerations result in outgoing connection strength values much
lower than one. The ratio of source-to-sink connectivity ($O:I$) for a target reef
represents the strength of the reef's outgoing larval flow relative to it's incoming larval
flow. This is calculated as: $O:I = \frac{S_O/N_O}{S_I/N_I}$. Where S represents the sum of
connectivity pobabilities, N represents the number of connections, O denotes outgoing
connectivity from the target reef and I denotes incoming connectivity to the target reef. If
a reef has no incoming connections, the $O:I$ equals the $S_O/N_O$ value. Incoming
connections in this calculation, do not directly represent the larval supply gained as the
breeding populations from which connections are coming from are not represented. A
source-to-sink ratio greater than 1 (a value greater than 0 on a log scale) indicates a reef
has greater outgoing than incoming larval flow. Reef larval retention probability indicates
the proportion of spawned larvae from a target reef that remain to settlement.
